Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-12-03 response_to_reviewers.txt

We thank the reviewers for their suggestions.

We are especially grateful to Reviewer 2, who may have offered the most detailed and constructive review we have ever received.

@@

Page 16
Plots of Figure 6:
1. Except for the axis labels, the plots have the similar problems as those of Figure 5.
2. Plots (c) and (d) are especially problematic w.r.t. overlaps.
3. Adding brightness (blue vs. light blue) as another distinctive feature (besides saturation) can
also hamper the readability (when the differences are as small as they are here).

45-56:

1. Which cases (of the four cases (a) – (d)) are meant with 'In both cases'?
* This referred to both of the cases mentioned in text (early and later access times). We revised the phrase to "these plots" to avoid confusion.

**** TODO: EM modified the below, though you can probably do better. Note that the text hasn't been modified accordingly yet.

2. What is the rationale (and meaning) behind '[excluding] pairs that are simple inclusions of pairs of trees with a smaller number of taxa'?
* These plots demonstrate the effect of curvature on access times when variables such as degree are controlled. Tree pairs that are inclusions of pairs of trees with a smaller number of taxa have largely unchanged access times to the smaller trees and clutter the figure.  For example, a pair of trees that both contain leaves 1 and 2 as siblings have very similar characteristics as the pair obtained by merging each sibling pair into a single leaf and relabeling the leaves. We rephrased this sentence to state "To limit clutter, these plots exclude pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g.\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings) as their access times distributions do not change substantially."

@@

1. Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n^2-4n-6) is not correct for all intended n. The reason is that for the ladder tree with three leaves, the small subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5) moves. Hence, a case distinction seems to be necessary here and for the inequality in the subsequent paragraph (lines 40-42)
* Thank you for pointing this out. To keep the lemma and proof simple (and because the n=3 tree space is trivial) we changed the assumption to n > 3, matching the subsequent lemmas.

2. The chain of (in)equalities in lines 40-42 does not run through, because adding (24-8n) leads to the n-free term '+12' (not '+16'). But this can be 'repaired' through the necessary case distinction n = 3 vs. n ≥ 4).
* As described above we restrict to n>3.

45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45 possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the typical notion of size in mind).
* We added the statement to line 45.

54-55: The statement 'a ladder tree has the greatest average moved subtree size' does not seem to hold for all n ≥ 4. For n = 4, the ladder tree has an average moved subtree size of 1, while it is 4/3 for the balanced tree. Similarly, the balanced tree has a larger one for n = 8 (1.55 vs. 1.39).  This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
* We should have been more clear on what constitutes a moved subtree here.  From our distribution of rSPR moves, the balanced tree with n = 4 leaves does not move any subtrees of size 2. There are moves of such trees that are equivalent to moves of leaves which appears to be the source of the confusion. As such, the average moved subtree size for n = 8 is 1.4 for the ladder tree and approximately 1.379 (160/116) for the balanced tree. We have added the following explanation to make this more clear to the reader:

### 2016-12-03 curvature.tex
Some small tweaks.

The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{felsenstein1990newick} representation of each tree as a string of characters (assuming integer labels and bounded reasonable values of $n$, see~\citep{whidden2016efficiently}).
These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
We construct a mapping from each tree $T_i$ to its order index in this ordering i$i$.
Begin with an empty graph $G$.
For each tree $T_i$, we first add a vertex $i$ to the graph and then use Corollary~\ref{cor:enumerate_neighbors} below to enumerate the $\OhOf{n^2}$ neighbors of $T_i$ in the rSPR graph in $\OhOf{n^3}$-time.
This efficient enumeration procedure is the key step required to achieve our desired running time of $\OhOf{mn^3}$.
We use the tree to index mappings to determine whether these trees are already vertices of the graph and, if so, add an edge in the graph from $T_i$ to each such neighbor $T_j$.
The high-level steps are as follows, and we show in Theorem~\ref{thm:construct_graph} that this algorithm is correct and can be implemented to run in the stated time.

@@

We consider the distribution of rSPR moves in terms of the number of nodes contained within the subtree to be moved.
Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
However, some of these moves will result in the same neighboring tree as other rSPR moves.
In particular, where we call the edge connecting the subtree rooted at that node to the rest of the tree the ``node's edge'', we have:
\begin{enumerate}
\item Moving $R$ to its sibling edge results in the same tree, not a neighboring tree,
\item Moving $R$ to its parent edge results in the same tree,
\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge, and
\item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
\end{enumerate}
We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is, every other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
To do so, we assign each of the neighbors to one node of the tree, such that any given neighbor can be obtained by moving the subtree rooted at its assigned node.
We assign $(2n-1-x)-2$ neighbors to children of the original non-$\rho$ root (lacking both an aunt and a grandparent), and $(2n-1-x)-4$ neighbors to each other non-root node.
Let $N(T,u)$ denote the neighbors of $T$ assigned to node $u$, obtained by moving the subtree $R$ rooted at $u$.
We thus achieve a new method for computing the neighborhood size:

@@

\begin{cor}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with trees in the neighborhood of $S$ such that $\dspr{T',S'} = 1$ for each $(T',S')$ pair, each neighbor of $T$ is paired with a single neighbor of $S$, and vice versa.
Note that a tree may be a neighbor of both $T$ and $S$ and thus involved in two different pairings.
\end{cor}
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
We consider the parent node of $R$ in both $T$ and $S$ to be the same source node for this purpose.
For each such $(T',S')$ pair, there are two cases, depending on whether $T'$ differs from $T$ by $R$ or another subtree.
If $T'$ and $T$ differ by a different subtree, then we can move $R$ in $T'$ or $S'$ to obtain the other member of the pair.
Thus, $\dspr{T',S'} = 1$.
If $T'$ and $T$ differ by $R$, however, then $T' = S'$.
Consider the set of such pairs $(T'_1, S'_1), (T'_2, S'_2), \ldots, (T'_p, S'_p)$.
Each tree of the pairs can be obtained from the others by moving $R$, so we simply pair $T'_1$ with $S'_2$, $T'_2$ with $S'_3$ and so on, finally pairing $T'_p$ with $S'_1$.
\end{proof}

@@

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $T$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Thus, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$.
	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

@@

To start we demonstrate the theorem in the case that $T_n$ and $S_n$ have the same number of neighbors.
First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
Let $(T'_n, S'_n)$ be any of the $\overlap$ pairs of neighbors of $(T_n, S_n)$ which are one rSPR move apart as per Corollary~\ref{cor:paired_neighbors}.
We pair these trees in the mass transport.
There are $\OhOf{kn}$ trees unmatched by this pairing, because $b$ and $a$ (in the nomenclature of Corollary~\ref{cor:paired_neighbors}) are bounded by $n$.
We can pair each of the unmatched trees arbitrarily with another tree of distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{T_n}$.

@@

Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
We also note that the lower bounds in this section do not follow from the more general setting described in \citep{Jost2013-ce}.
However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
This follows by assuming that each of the unpaired trees is moved the minimum distance of one, that is the curvature for this pair of trees falls between $-\mu$ and $-2\mu$.

>>>>

The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \citep{felsenstein1990newick} representation of each tree as a string of characters (assuming integer labels and bounded reasonable values of $n$, see~\citep{whidden2016efficiently}).
These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
We construct a mapping from each tree $T_i$ to its order index in this ordering $i$.
Begin with an empty graph $G$.
For each tree $T_i$, we first add a vertex $i$ to the graph and then use Corollary~\ref{cor:enumerate_neighbors} below to enumerate the $\OhOf{n^2}$ neighbors of $T_i$ in the rSPR graph in $\OhOf{n^3}$-time.
This efficient enumeration procedure is the key step required to achieve our desired running time of $\OhOf{mn^3}$.
We use the tree to index mappings to determine whether these trees are already vertices of the graph and, if so, add an edge in the graph from $T_i$ to each such neighbor $T_j$.
The high-level steps are as follows, and we show in Theorem~\ref{thm:construct_graph} that this algorithm is correct and can be implemented to run in the stated time.

@@

%EM "distribution" here and below seems strange to me because we aren't always sampling. Wouldn't "enumeration" work?
We consider the distribution of rSPR moves in terms of the number of nodes contained within the subtree to be moved.
Recall that a tree with $n$ leaves has $2n-1$ total nodes (ignoring the artificial $\rho$ node).
Given a subtree $R$ with $x$ nodes, observe that there are $2n-1 - x$ possible locations to regraft $R$.
However, some of these moves will result in the same neighboring tree as other rSPR moves.
In particular, where we call the edge connecting the subtree rooted at that node to the rest of the tree the ``node's edge'', we have:
\begin{enumerate}
\item Moving $R$ to its sibling edge results in the same tree, not a neighboring tree,
\item Moving $R$ to its parent edge results in the same tree,
\item Moving $R$ to its grandparent edge is the same as moving its aunt to its sibling edge, and
\item Moving $R$ to its aunt edge is the same as moving its aunt to $R$'s edge.
\end{enumerate}
We prove in Lemma~\ref{lem:compute_degree} that this list is exhaustive, that is, every other pair of $R$ and destination edge $e$ results in a unique rSPR neighbor.
To do so, we assign each of the neighbors to one node of the tree, such that any given neighbor can be obtained by moving the subtree rooted at its assigned node.
We assign $(2n-1-x)-2$ neighbors to children of the original non-$\rho$ root (lacking both an aunt and a grandparent), and $(2n-1-x)-4$ neighbors to each other non-root node.
Let $N(T,u)$ denote the neighbors of $T$ assigned to node $u$, obtained by moving the subtree $R$ rooted at $u$.
We thus achieve a new method for computing the neighborhood size:

@@

\begin{cor}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$\overlap := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with trees in the neighborhood of $S$ such that $\dspr{T',S'} = 1$ for each $(T',S')$ pair, each neighbor of $T$ is paired with a single neighbor of $S$, and vice versa.
Note that a tree may be a neighbor of both $T$ and $S$ and thus involved in two different pairings.
\end{cor}
\begin{proof}
%EM Unlike you and the reviewer, I can't figure out what source and target mean here-- are they defined?
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $\overlap$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
We consider the parent node of $R$ in both $T$ and $S$ to be the same source node for this purpose.
For each such $(T',S')$ pair, there are two cases, depending on whether $T'$ differs from $T$ by $R$ or another subtree.
If $T'$ and $T$ differ by a different subtree, then we can move $R$ in $T'$ or $S'$ to obtain the other member of the pair.
Thus, $\dspr{T',S'} = 1$.
If $T'$ and $T$ differ by $R$, however, then $T' = S'$.
Consider the set of such pairs $(T'_1, S'_1), (T'_2, S'_2), \ldots, (T'_p, S'_p)$.
Each tree of the pairs can be obtained from the others by moving $R$, so we simply pair $T'_1$ with $S'_2$, $T'_2$ with $S'_3$ and so on, finally pairing $T'_p$ with $S'_1$.
\end{proof}

@@

\begin{lem}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lem}
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
    %EM Don't you mean displaces $R$ from its original position?
	In the latter case, $S$ is reached from $U$ by moving $Q$ in a manner that displaces $T$ from its original position, because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Thus, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$, where $L(Q)$ is the label set of $Q$.
	Then leaves $r' \in R$, $q' \in Q$, and $v' \in V$, for some subtree $V$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}

@@

To start we demonstrate the theorem in the case that $T_n$ and $S_n$ have the same number of neighbors.
First we claim that $W_{1,n}$ is bounded above by $(\degree{T_n}+\OhOf{kn})/\degree{T_n}$ by exhibiting a mass transport program satisfying that bound.
Let $(T'_n, S'_n)$ be any of the $\overlap$ pairs of neighbors of $(T_n, S_n)$ which are one rSPR move apart as per Corollary~\ref{cor:paired_neighbors}.
We pair these trees in the mass transport.
%EM Sorry, I don't see b and a in cor:paired_neighbors.
There are $\OhOf{kn}$ trees unmatched by this pairing, because $b$ and $a$ (in the nomenclature of Corollary~\ref{cor:paired_neighbors}) are bounded by $n$.
We can pair each of the unmatched trees arbitrarily with another tree of distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{T_n}$.

@@

Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
We also note that the lower bounds in this section do not follow from the more general setting described in \citep{Jost2013-ce}.
However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
This follows by assuming that each of the unpaired trees is moved the minimum distance of one, that is, the curvature for this pair of trees falls between $-\mu$ and $-2\mu$.
